Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-02 paper/paper_no_event.tex
Paper: small corrections, in particular change sign of A1

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $M_{A,B}$ are the masses of bodies $A$ and $B$, $M = M_A + M_B$ is the total mass, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.

@@

Tidal effects also enter the radiative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu} as
\be
	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
\ee
The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived. The Hamiltonian equations of motion with this radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}. The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics. We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.

@@

After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below. This reduces the number of matter parameters from ten to two.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform,
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
\end{equation}
where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_{0} + \phi_{0} + \frac{\pi}{4} - \frac{3}{128\eta}v^{-5} \left[\Phi_{\rm F2}^{\rm PP}(\eta) \right. \nonumber\\
	& \left. \qquad + \Phi_{\rm F2}^{\rm Spin} (\eta, \chi_1, \chi_2) + \Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right] \,,
\end{align}
where $t_{0}$ and $\phi_{0}$ are constants related to the freedom of choosing the time and phase of coalescence. We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$. We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet:2013haa}).

The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $. We include spin-orbit terms up to 3.5PN~\cite{Blanchet:2013haa}. For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{subsec:QM}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms. The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.

>>>>

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $m_{A,B}$ are the masses of bodies $A$ and $B$, $M = m_A + m_B$ is the total mass, $X_{A,B}=m_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=m_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/m_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.

@@

Tidal effects also enter the radiative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
%\be
%	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
%\ee

The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived. The Hamiltonian equations of motion with this radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}. The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics. We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.

@@

After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=m_B/m_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below. This reduces the number of matter parameters from ten to two.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform,
\begin{equation}
A_{\rm F2} = -\sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
\end{equation}
where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The sign is set following the conventions of \texttt{LAL}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_{0} + \phi_{0} + \frac{\pi}{4} - \frac{3}{128\eta}v^{-5} \left[\Phi_{\rm F2}^{\rm PP}(\eta) \right. \nonumber\\
	& \left. \qquad + \Phi_{\rm F2}^{\rm Spin} (\eta, \chi_1, \chi_2) + \Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right] \,,
\end{align}
where $t_{0}$ and $\phi_{0}$ are constants related to the freedom of choosing the time and phase of coalescence. We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$. We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet2014Review}).

The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $. We include spin-orbit terms up to 3.5PN~\cite{Blanchet2014Review}. For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{subsec:QM}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms. The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.
